Synthesising Complex Termination Arguments

### 2014-06-23 papers/termination/paper.tex
Flesh out 2nd order termination

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{pifont}
\usepackage{listings}
\usepackage{graphicx}
\usepackage{xspace}
\usepackage{pgf}
\usepackage[noend]{algpseudocode}
\usepackage{algorithm}
\usepackage{multicol}
\usepackage{appendix}
\usepackage{caption,subcaption}
\DeclareCaptionType{copyrightbox}
\usepackage{subfig}
\usepackage{multirow}
\usepackage{tikz}
\usetikzlibrary{arrows, automata, shapes}

@@

\section{Termination as Second-Order Satisfaction}
In this section we will show how several variations of the halting problem can be defined in second-order
logic.  As it turns out, all of these variations can be defined in a fragment of second-order logic
we will refer to as the \emph{synthesis fragment}.

\begin{figure*}
\begin{definition}[Second-order Termination Formula]
\label{def:termination-formula}
\begin{align*}
 \exists R \cdot \forall x, x' . & R(x) > 0 ~ \wedge \\
                             & T(x, x') \rightarrow R(x) > R(x')
\end{align*}
\end{definition}

\begin{definition}[Conditional Termination Formula]
 \begin{align*}
  \exists R, I \cdot \forall x, x' . & I(x) \rightarrow V(x) ~ \wedge \\
                                 & L(x) \wedge V(x) \wedge T(x, x') \rightarrow V(x') \wedge R(x) > 0 \wedge R(x) > R(x')
 \end{align*}
\end{definition}

\begin{definition}[Multiple Loops Termination Formula]
 \begin{align*}
  \exists R, V_1, V_2 \cdot \forall x, x' . & I(x) \rightarrow V_1(x) ~ \wedge \\
                                        & V_1(x) \wedge L_1(x) \wedge T(x, x') \rightarrow V_1(x') ~ \wedge \\
                                        & V_1(x) \wedge \lnot L_1(x) \rightarrow V_2(x) ~ \wedge \\
                                        & V_2(x) \wedge L_2(x) \wedge T(x, x') \rightarrow V_2(x') \wedge R(x) > 0 \wedge R(x) > R(x')
 \end{align*}
\end{definition}

@@

\begin{definition}[Deterministic Non-Terminaton Formula]
 \begin{align*}
  \exists N, x_0 \cdot \forall x, x' \cdot & I(x_0) \wedge L(x_0) \wedge N(x_0) ~ \wedge \\
							& N(x) \wedge L(x) \wedge T(x, x') \rightarrow N(x') \wedge L(x')
 \end{align*} 
\end{definition}

\caption{The halting problem defined in second-order logic.\label{fig:termination-formulae}}
\end{figure*}

@@

>>>>

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{pifont}
\usepackage{listings}
\usepackage{graphicx}
\usepackage{xspace}
\usepackage{pgf}
\usepackage[noend]{algpseudocode}
\usepackage{algorithm}
\usepackage{multicol}
\usepackage{appendix}
\usepackage{caption,subcaption}
\DeclareCaptionType{copyrightbox}
\usepackage{subfig}
\usepackage{multirow}
\usepackage{framed}
\usepackage{tikz}
\usetikzlibrary{arrows, automata, shapes}

@@

\section{Termination as Second-Order Satisfaction}
In this section we will show how several variations of the halting problem can be defined in second-order
logic.  As it turns out, all of these variations can be defined in a fragment of second-order logic
we will refer to as the \emph{synthesis fragment}.  We will make use of this observation to frame the halting
problem as a program synthesis problem, allowing us to employ the program synthesiser described in
Section~\ref{sec:synthesis}.

\begin{figure*}
\begin{framed}
\begin{definition}[Second-order Termination Formula]
\label{def:termination-formula}
\begin{align*}
 \exists R \cdot \forall x, x' \cdot & L(x) \wedge T(x, x') \rightarrow R(x) > 0 \wedge R(x) > R(x')
\end{align*}
\end{definition}

\begin{definition}[Conditional Termination Formula]
\label{def:conditional-termination-formula}
 \begin{align*}
  \exists R, V \cdot \forall x, x' \cdot & I(x) \rightarrow V(x) ~ \wedge \\
                                 & L(x) \wedge V(x) \wedge T(x, x') \rightarrow V(x') \wedge R(x) > 0 \wedge R(x) > R(x')
 \end{align*}
\end{definition}

\begin{definition}[Multiple Loops Termination Formula]
\label{def:multi-termination-formula}
 \begin{align*}
  \exists R, V_1, V_2 \cdot \forall x, x' \cdot & I(x) \rightarrow V_1(x) ~ \wedge \\
                                        & V_1(x) \wedge L_1(x) \wedge T(x, x') \rightarrow V_1(x') ~ \wedge \\
                                        & V_1(x) \wedge \lnot L_1(x) \rightarrow V_2(x) ~ \wedge \\
                                        & V_2(x) \wedge L_2(x) \wedge T(x, x') \rightarrow V_2(x') \wedge R(x) > 0 \wedge R(x) > R(x')
 \end{align*}
\end{definition}

@@

\begin{definition}[Deterministic Non-Termination Formula]
\label{def:deterministic-nonterm-formula}
 \begin{align*}
  \exists N, x_0 \cdot \forall x, x' \cdot & I(x_0) \wedge L(x_0) \wedge N(x_0) ~ \wedge \\
							& N(x) \wedge L(x) \wedge T(x, x') \rightarrow N(x') \wedge L(x')
 \end{align*} 
\end{definition}
\end{framed}
\end{figure*}

@@

The existence of a ranking function for a single loop is equivalent to the satisfiability
of the formula from Definition~\ref{def:termination-formula}.  If this formula is satisfiable,
$R$ is a ranking function that proves the loop $L$ terminates, so the termination of $L$
is equivalent to the satisfiability of Definition~\ref{def:termination-formula}.

A slightly more involved case occurs if $L$'s termination depends on the state it begins
executing in, i.e. $L$ is \emph{conditionally terminating}.  This property is equivalent to the
formula of Defintion~\ref{def:conditional-termination-formula}.  If this formula is satisfiable,
$V$ is an inductive invariant of $L$ that is established by the initial states $I$.
$R$ is then a ranking function for $L$ as restricted by $V$ -- that is to say, $R$ need only
be well founded on those states satisfying $V$.  But since $V$ is an inductive invariant of $L$,
$R$ is strong enough to show that $L$ terminates from any of its initial states.  We say
that $V$ is a \emph{termination invariant} for $L$ and that $R$ proves termination relative to $V$.

The conditional termination formula becomes more complex in the presence of multiple loops.
Consider the following program with two loops $L_1$ and $L_2$:

\begin{lstlisting}
while (L1) { ... }
while (L2) { ... }
\end{lstlisting}

The termination problem for $L_2$ is equivalent to the satisfiability of
Definition~\ref{def:multi-termination-formula}.  If this formula is satisfiable,
$V_1$ is an inductive invariant of $L_1$ and $V_2$ is a termination invariant of $L_2$.
$V_1$ is established by the initial conditions $I$ and is strong enough to establish
$V_2$ when conjoined with the negation of $L_1$'s guard.  As before, $R$ is a ranking
function relative to $V_2$.

Non-termination of the loop $L$ is encoded by Definition~\ref{def:nonterm-formula}.
If this formula is satisfiable, $N$ is a recurrence set for $L$ which proves
$L$'s non-termination.  However this formula has one more level of quantifier alternation
than the others (it is an $\exists \forall \exists$ formula), which will cause us problems
shortly.  We observe that in the case that $T$ is deterministic (that is, each state has
exactly one successor),  the non-termination problem is equivalent to the
satisfiability of Definition~\ref{def:deterministic-nonterm-formula}.
